Databases are an effective and efficient method for storage and access of mass-spectrometry data

### 2025-07-18 manuscript/manuscript.Rmd
ce from thesis committee

We chose to focus on liquid-chromatography mass-spectrometry (LC-MS) data given its widespread use and fundamentally simple raw data structure as tuples consisting of retention time, *m/z*, and intensity, though the tidy framework here can be extended easily to other MS data (Supplemental Figure 1). We performed a literature search for mass-spectrometry data formats that have been published in the last 15 years and attempted to find or construct parsers for each format in Python, a popular high-level interpreted language. Each parser was written to perform three common exploratory data analysis operations on full-scan data and three common operations on MS/MS fragmentation data. Full scan queries consisted of 1) single scan extraction by scan number, 2) retention time range extraction of all scans within a specified retention time range, and 3) chromatogram extraction, which collects the ions within a specified parts-per-million (PPM) error of a known mass. These queries generally correspond to the methods used in Bouyssié et al. (2015) @Bouyssie2015, which performed similar tests benchmarking the mzDB format against mz5 and an mzML parser. Note that the chromatogram extraction does not extract a precompiled chromatogram of the sort commonly found at the end of mzML files or as a result of SIM/PRM analysis but instead refers to sifting through the raw data for data tuples with an *m/z* value between specified bounds. MS/MS queries involved extracting three relevant subsets, consisting of 1) a single scan extraction by scan number similar to that of the full scan, 2) extraction of all the fragments associated with a precursor *m/z* within a given PPM, and 3) extraction of all fragments with *m/z* values within a given PPM.

@@

Timing data was obtained on an Intel Xeon CPU with two X5650 (\@2.67 GHz) processors and 24 total cores running Windows 10 Pro (64 bit version). 96 gigabytes of RAM (DDR3 @ 1333 MHz) were available and a solid-state drive was used for disk storage.

@@

The simplest and most abundantly documented query was the extraction of a single spectrum. In many ways, this is the fundamental unit of mass spectrometry and thus many formats are highly optimized for its extraction into manipulatable data (Figure 2A and 2D). Here, we found that the mzML and mzMLb file types were consistently the slowest to parse and required multiple seconds, likely highlighting inefficiencies in the `pyteomics` package used to parse both file types. `pyopenms` also struggled to open and extract a specific scan, requiring several seconds due in large part to the expensive initiation function, after which requests were orders of magnitude faster (Supplemental Figure 2). It is also worth noting that while both of these packages provided rapid extraction of a *random* spectrum, a significant overhead was introduced by needing to scan through the file to find a *specific* spectrum by scan number. Scans are not always consecutive and no metadata was obviously available that would have allowed using the index directly to a specific scan number.

@@

We were also unable to test several other recent and promising formats. Aird does not report full query times for any of the metrics reviewed here, though they claim their StackZDPD algorithm [@Wang2022] can improve decompression speed by three times and that the file size is 54% of the vendor file.[@Lu2022] Similarly, the toffee format for time-of-flight DIA data reports sizes about equal to vendor or 60% of centroid mzml + numpress with query speeds 4 times faster for scans (spectrum-centric, 2 seconds for mzML, 0.5 for toffee) and 100 times faster for chromatograms (peptide-centric, 168 seconds for mzML, 1.8 for toffee).[@Tully2020] The Unified Ion Mobility Format (UIMF) @Beagley2009 from Pacific Northwest National Laboratory format did not report direct comparisons to any of the available formats and thus we must remain unclear on its performance capabilities.

>>>>

We chose to focus on liquid-chromatography mass-spectrometry (LC-MS) data given its widespread use and fundamentally simple raw data structure with entries consisting of retention time, *m/z*, and intensity, though the tidy framework here can be extended easily to other MS data (Supplemental Figure 1). We performed a literature search for mass-spectrometry data formats that have been published in the last 15 years and attempted to find or construct parsers for each format in Python, a popular high-level interpreted language. Each parser was written to perform three common exploratory data analysis operations on full-scan data and three common operations on MS/MS fragmentation data. Full scan queries consisted of 1) single scan extraction by scan number, 2) retention time range extraction of all scans within a specified retention time range, and 3) chromatogram extraction, which collects the ions within a specified parts-per-million (PPM) error of a known mass. These queries generally correspond to the methods used in Bouyssié et al. (2015) @Bouyssie2015, which performed similar tests benchmarking the mzDB format against mz5 and an mzML parser. Note that the chromatogram extraction does not extract a precompiled chromatogram of the sort commonly found at the end of mzML files or as a result of SIM/PRM analysis but instead refers to sifting through the raw data for data tuples with an *m/z* value between specified bounds. MS/MS queries involved extracting three relevant subsets, consisting of 1) a single scan extraction by scan number similar to that of the full scan, 2) extraction of all the fragments associated with a precursor *m/z* within a given PPM, and 3) extraction of all fragments with *m/z* values within a given PPM.

@@

Timing data was obtained on an Intel Xeon CPU with two X5650 (\@2.67 GHz) processors and 24 total cores running Windows 10 Pro (64-bit version). 96 gigabytes of RAM (DDR3 @ 1333 MHz) were available and a solid-state drive was used for disk storage.

@@

The simplest and most abundantly documented query was the extraction of a single spectrum. In many ways, this is the fundamental unit of mass spectrometry and thus many formats are highly optimized for its extraction into manipulable data (Figure 2A and 2D). Here, we found that the mzML and mzMLb file types were consistently the slowest to parse and required multiple seconds, likely highlighting inefficiencies in the `pyteomics` package used to parse both file types. `pyopenms` also struggled to open and extract a specific scan, requiring several seconds due in large part to the expensive initiation function, after which requests were orders of magnitude faster (Supplemental Figure 2). It is also worth noting that while both of these packages provided rapid extraction of a *random* spectrum, a significant overhead was introduced by needing to scan through the file to find a *specific* spectrum by scan number. Scans are not always consecutive and no metadata was obviously available that would have allowed using the index directly to a specific scan number.

@@

We were also unable to test several other recent and promising formats. Aird does not report full query times for any of the metrics reviewed here, though they claim their StackZDPD algorithm [@Wang2022] can improve decompression speed by three times and that the file size is 54% of the vendor file.[@Lu2022] Similarly, the toffee format for time-of-flight DIA data reports sizes about equal to vendor or 60% of centroid mzML + numpress with query speeds 4 times faster for scans (spectrum-centric, 2 seconds for mzML, 0.5 for toffee) and 100 times faster for chromatograms (peptide-centric, 168 seconds for mzML, 1.8 for toffee).[@Tully2020] The Unified Ion Mobility Format (UIMF) @Beagley2009 from Pacific Northwest National Laboratory format did not report direct comparisons to any of the available formats and thus we must remain unclear on its performance capabilities.
